---
title: "Control Flow"
---

## Indexing

In the @sec-r-objects, we discussed about different types of R objects. For example, a vector can be a certain data type with a set number of elements. Here we construct a vector called `x` increasing from -5 to 5 by one unit:

```{r}
(x <- -5:5)
```

The vector `x` has 11 elements. If I want to know what the 6th element of `x`, I can index the 6th element from a vector. To do this, we use `[]` square brackets on `x` to index it. For example, we index the 6th element of `x`:

```{r}
x[6]
```
When ever we use `[]` next to an R object, it will print out the data to a specific value inside the square brackets. We can index an R object with multiple values:

```{r}
x[1:3]
x[c(3,9)]
```
Notice how the second line uses the `c()`. This is necessary when we want to specify non-contiguous elements. Now let's see how we can index a matrix

## If/Else Statements

## for Loops


## while Loops


```{r}
# Control Flow: if else statements
x <- rnorm(1)

## Logical Statements

x > 0 

## if statements

if (x > 0) {
  print("Positive")
} 

## else statements

if (x > 0){
  print("Positive")
} else {
  print("Non-Positive")
}

## Example 2
y <- rnorm(1)


if (y > 0){
  print("Positive")
  print(y)
  mean(y)
} else {
  print("Non-Positive")
  print(y)
  length(y)
}

```

```{r}
# Control Flow: else if statements
(x <- sample(-1:1,1))

## Logical Statements

x > 0 

## if statements

if (x > 0) {
  print("Positive")
} 

## else if statements

if (x > 0) {
  print("Positive")
} else if (x < 0) {
  print("Negative")
} else {
  print("Zero")
}


## Example 2
y <- sample(-1:1,1)


if (y > 0){
  print("Positive")
} else if (y < 0) {
  print("Negative")
} else {
  print("Zero")  
}

```



```{r}
# Control Flow: break & next function -----

## Function
err_fx <- function(x){
  if (x>0){
    return(x)
  } else {
    stop("x is not positive")
  }
}

(y <- rnorm(1))
# err_fx(y)

## Loop Example
x <- rnorm(100)
loop <- c()
for (i in seq_along(x)) {
  try_err <- try(err_fx(x[i]), silent = T)
  if (inherits(try_err, "try-error")){
    loop[i] <- 0
  } else {
    loop[i] <- try_err
  }
}


## Break -----
x <- rnorm(100)
loop <- c()
for (i in seq_along(x)) {
  try_err <- try(err_fx(x[i]), silent = T)
  if (inherits(try_err, "try-error")){
    break
  } else {
    loop[i] <- try_err
  }
}

## Next -----
x <- rnorm(100)
loop <- c()
for (i in seq_along(x)) {
  try_err <- try(err_fx(x[i]), silent = T)
  if (inherits(try_err, "try-error")){
    next
  } else {
    loop[i] <- try_err
  }
}

x <- rnorm(100)
loop <- c()
for (i in seq_along(x)) {
  try_err <- try(err_fx(x[i]), silent = T)
  if (inherits(try_err, "try-error")){
    next
  } else {
    loop <- c(loop, try_err)
  }
}

```


```{r}
# Control Flow: try function -----

## Function
err_fx <- function(x){
  if (x>0){
    return(x)
  } else {
    stop("x is not positive")
  }
}

## Example -----
(y <- rnorm(1))
# err_fx(y)

## try function ----

y_err <- try(err_fx(y), silent = T)

## Example ----
# x <- rnorm(100)
# loop <- c()
# for (i in x){
#   loop[i] <- err_fx(i)
# }

## Using try
x <- rnorm(100)
loop <- c()
for (i in seq_along(x)) {
  try_err <- try(err_fx(x[i]), silent = T)
  if (inherits(try_err, "try-error")){
    loop[i] <- 0
  } else {
    loop[i] <- try_err
  }
}


```

```{r}
# Control Flow: Loops ----

# Loops are used to conduct repetitive/iterative tasks
# Each iteration conducts a task given a set of values
# The values for each iteration change as the loop moves
# from one iteration to another

## for Anatomy -----

# for (i in vector) {
#   Perform Task
# }

## Printing Example ------

### print number 1 through 5, separately

# We want to do this:
print(1); print(2); print(3); print(4); print(5)

# We don't want this:
print(1:5)

### Using a loop
for (i in 1:5){
  print(i)
}

## Printing Letters -----
### Print all the letters, seperately
print(letters)

for (i in 1:26){
  print(letters[i])
}

## cleaner
for (i in seq_along(letters)){
  print(letters[i])
}

## cleanest

for (i in letters){
  print(i)
}
```

```{r}
# Control Flow: Nested Loops ----
library(greekLetters)
letters_new <- letters[1:3]
greek_lower <- greek_vector[1:24]

paste(letters_new[1], greek_lower[1], sep = "")
paste(letters_new[1], greek_lower[2], sep = "")

paste(letters_new[2], greek_lower[1], sep = "")
paste(letters_new[2], greek_lower[2], sep = "")

paste(letters_new[3], greek_lower[1], sep = "")
paste(letters_new[3], greek_lower[2], sep = "")


## Inefficient way
for(i in greek_lower){
  print(paste(letters_new[1], i, sep = ""))
}

for(i in greek_lower){
  print(paste(letters_new[2], i, sep = ""))
}

for(i in greek_lower){
  print(paste(letters_new[3], i, sep = ""))
}

for (i in 1:3){
  for (ii in greek_lower){
    print(paste(letters_new[i], ii, sep = ""))
  }
}

for (i in letters_new){
  for (ii in greek_lower){
    print(paste(i, ii, sep = ""))
  }
}

# for (i in vector) {
#   for (ii in vector) {
#     for (iii in vector) {
#       for (iiii in vector) {
#         
#       }
#     }
#   }
# }


```


```{r}
# Control Flow: Loops ----

# for (i in vector) {
#   Perform Task
# }

## Vector construction -----
# x²
### Method 1
x <- rnorm(1000)
x^2
output <- c()
for (i in seq_along(x)){
  output <- c(output, x[i]^2)
}

### Method 2
x <- rnorm(1000)
output <- c()
for (i in seq_along(x)){
  output[i] <- x[i]^2
}

### Method 3
x <- rnorm(1000)
output <- vector(length = length(x))
for (i in seq_along(x)){
  output[i] <- x[i]^2
}

```



```{r}
# Control Flow: While Loops -----
x <- -5:5
plot(x, exp(x), type = "l")

## Asymptotic -----
abs(exp(-14)-exp(-13))

## while statements -----
# while (condition) {
#   
# }


## while loops ----
x <- 1
diff <- 1
while (diff > 1e-20) {
  old_x <- x
  x <- x - 1
  diff <- abs(exp(x) - exp(old_x))
}
print(x)
print(diff)


```

```{r}
# Control Flow: Infinite While Loops -----
(y <- rnorm(1))
z <- as.integer(2)

# logical operator &
is.integer(z) & z > 0
is.integer(y) & y > 0

x <- -50:50
plot(x, sin(x), type = "l")

## Asymptotic -----
abs(exp(13)-exp(12))

## inf while loops ----
x <- 1
diff <- 1
# while (diff > 1e-20) {
#   old_x <- x
#   x <- x + 1
#   diff <- abs(exp(x) - exp(old_x))
# }
print(x)
print(diff)

# x <- 1
# diff <- 1
# while (diff > 1e-20) {
#   old_x <- x
#   x <- x + 1
#   diff <- abs(sin(x) - sin(old_x))
# }
# print(x)
# print(diff)




## while loops ----
x <- 1
counter <- 0
diff <- 1
while (diff > 1e-20 & counter < 30) {
  old_x <- x
  x <- x + 1
  diff <- abs(exp(x) - exp(old_x))
  counter <- counter + 1
}
print(x)
print(diff)
print(counter)

x <- 1
counter <- 0
diff <- 1
while (diff > 1e-20 & counter < 10^3) {
  old_x <- x
  x <- x + 1
  diff <- abs(sin(x) - sin(old_x))
  counter <- counter + 1
}
print(x)
print(diff)
print(counter)

```

