---
title: "Quarto Documents"
knitr:
  opts_chunk: 
    echo: true
    eval: false
    tidy: true
    tbl-cap-location: bottom
    tidy.opts:
      width.cutoff: 60
    R.options:
      digits: 3
bibliography: document.bib      
---


## Introduction

[Quarto](https://quarto.org/) is a file type used to create technical reports while including both R code, or other programming languages, and output in a document. A qmd^[QMD is the file extension to use the Quarto engine. For this document Quarto and QMD are used interchangeably.] file is a fancy R Script containing extra capabilities. Additionally, qmd files allow for citations, footnotes, mathematical expressions, links, and much more. Once the document is finished, it can be rendered to a word file, pdf, html file, and much more. Quarto is the considered the next generation of [RMarkdown](https://rmarkdown.rstudio.com/).

## Anatomy of a Quarto Document

There are three main components in an qmd file: the YAML header, R code, and basic text.

The YAML header contains information on how to render the document. It is located at the beginning of the document surrounded by 3 dashes (`---`) above and below it. For starters, the YAML header will contain a 'title', 'author', 'date', and 'output' line. 



The R code is located in a block known as chunks. A chunk tells RStudio to read the next lines as code. A chunk begins with three back ticks followed by `{r}` and ends with three back ticks. Everything in between the back ticks will be executed by R. In RStudio, a chunk can be inserted using the keyboard shortcut `ctrl+alt+I` or `cmd+option+I`. 

An example of an R chunk is shown below:

```{r}
#| echo: fenced
mean(mtcars$mpg)
```

The R chunk will be rendered as below:

```{r}
#| eval: false
mean(mtcars$mpg)
```


Notice the chunk includes the code in a block followed by the output from the console.

The last component of an qmd document is the text. Write anywhere in the document, and it will be rendered as is.

## Chunk Options

R chunks have options that will alter how the code or the output is rendered. The chunk options can be set either globally to affect the entire qmd document or locally to affect only an individual chunk. For more information about chunk options, visit <https://yihui.org/knitr/options/>

### Global Chunk Options

To set global chunks options, add the two lines to YAML header:

```{default}
#| echo: true
---
knitr:
  opts_chunk: 
---
```

Followed by the chunks and R options you want to set:

```{default}
#| echo: true
---
knitr:
  opts_chunk:
    eval: false
    tidy: styler
    R.options:
      digits: 2
---

```




A couple of recommended chunk options set globally are `eval: false`, and `tidy: styler`. These options make rendering the document easier. 

One chunk option is `tidy: styler`. This tells Quarto to prevent code from printing in a long line, possibly off the page. For example, look at the output of this chunk:

```{r}
#| tidy: false
## This comment is designed to show what happens when all your code is in 1 line. This is fine when you are coding, but when you are putting it in a report, it will run off the page.
```

Notice the comment being printed off the page. Using the options `tidy: styler`, the chunk is rendered as

```{r}
## This comment is designed to show what happens when all your code is in 1 line. This is fine when you are coding, but when you are putting it in a report, it will run off the page.
```


The last 2 lines control how R will compute and print output. `R.Options:` tells Quarto that the options R will be changed, and each line after alters options. `digits: 2` indicates R to use 2 [significant digits](https://en.wikipedia.org/wiki/Significant_figures). 

### Local Chunk Options

Local chunk options can be used to control an individual chunk will behave. To control a specific chunk, place the option below the `{r}` identifier and use the `#|` chunk option indicator. An example is povided below:

```{r}
#| echo: fenced
#| eval: false
#| tidy: false

mean(mtcars$mpg)
```

The chunk option `eval` set to `false` tells Quarto to not evaluate the code within the chunk. Notice how the output was not printed the R chunk above. When we set `eval` to `true`, the output is printed:

```{r}
#| echo: fenced 
#| eval: true
mean(mtcars$mpg)
```


The `echo` option will control if the code within the chunk should be printed in the document. This next chunk contains `#| echo: true`:

```{r}
#| echo: true
#| eval: true
mean(mtcars$mpg)
```

Now the chunk contains `#| echo: false`:

```{r}
#| echo: false
#| eval: true
mean(mtcars$mpg)
```

The R Code disappears.

There are chunk options for figures as well. A few options are `fig-height`, `fig-width`, `fig-align`, and `fig-cap`.

This chunk contains `fig-height: 3.5; fig-width: 3.5; fig-align: left`.

```{r}
#| echo: fenced
#| eval: true
#| fig-height: 3.5
#| fig-width: 3.5
#| fig-align: left

plot(mtcars$mpg,mtcars$drat)
```

The chunk options tells RStudio to create an image that is 3.5 inches in height and width, and align the image to the left.

The following chunk contains `fig-height: 3.5; fig-width: 3.5; fig-align: left; fig-cap: "This is a scatter plot of MTCARS' MPG and DRAT"; label: fig-mtcars`.

```{r}
#| echo: fenced
#| eval: true
#| fig-height: 3.5
#| fig-width: 3.5
#| fig-align: center
#| fig-cap: "This is a scatter plot of MTCARS' MPG and DRAT"
#| label: fig-mtcars

plot(mtcars$mpg,mtcars$drat)
```

The chunk adds a caption with `fig-cap` and reference label with `label`. The label of the plot can be referenced later in the document. @fig-mtcars can be referenced with `@fig-mtcars`.

### Inline Code

Instead of evaluating code in a chunk, code can evaluated in the text instead. For example, if we want to write the mean mpg in `mtcars` is `{r} mean(mtcars$mpg)`, one can type ``{r} mean(mtcars$mpg)``, surrounded by back ticks (`` ` ``), instead of writing the entire number and risk of miscopying the results.

## Formatting

Qmd files contain basic formatting capabilities. The use of the `#` followed by text creates a heading. Using two or more `#` symbols will create subheadings based on the number of `#`. A text is *italicized* by surrounding the text with one asterisk (`*italicized*`). A text is **boldfaced** by surrounding it with 2 asterisk (`**boldfaced**`).

To create an unordered list, use the `-` symbol at the beginning of each line. To create a sub-item, press the tab button twice (4 spaces), then the `-` symbol. Repeat this method for further sub-items.

-   First Item

-   Second Item

    -   First Sub-Item

        -   First Sub-Sub-Item

            -   First Sub-Sub-Sub-Item

To created an ordered list, type the number followed by a period for each line. To create sub-lists, press the tab button twice and order them appropriately.

1.  First

2.  Second

    a.  First

    b.  Second

        1)  First
        2)  Second

> A block quote is created with the `>` symbol at the beginning of a line.

Qmd files allows a table to be constructed in 2 ways, manually or using a package such as the `gt` package. A table is manually created by using `|`, `:`, and `-`. The first line contains `|` and the column names in between. The second line contains `|:-|:-|` which indicates how the table is aligned. The location of `:` symbol just tells RStudio about the alignment. Below is the example code of @tbl-greektable:

```{default}
| Letter     |   Lowercase   |      Code     |  Uppercase |    Code    |
|:-----------|:-------------:|:-------------:|:----------:|:----------:|
| alpha      |    $\alpha$   |    `\alpha`   |     --     |     --     |
| beta       |    $\beta$    |    `\beta`    |     --     |     --     |
| gamma      |    $\gamma$   |    `\gamma`   |  $\Gamma$  |  `\Gamma`  |
| delta      |    $\delta$   |    `\delta`   |  $\Delta$  |  `\Delta`  |
| epsilon    |   $\epsilon$  |   `\epsilon`  |     --     |     --     |
| zeta       |    $\zeta$    |    `\zeta`    |     --     |     --     |
| eta        |     $\eta$    |     `\eta`    |     --     |     --     |
| theta      |    $\theta$   |    `\theta`   |  $\Theta$  |  `\Theta`  |
| iota       |    $\iota$    |    `\iota`    |     --     |     --     |
| kappa      |    $\kappa$   |    `\kappa`   |     --     |     --     |
| lambda     |   $\lambda$   |   `\lambda`   |  $\Lambda$ |  `\Lambda` |
| mu         |     $\mu$     |     `\mu`     |     --     |     --     |
| nu         |     $\nu$     |     `\nu`     |     --     |     --     |
| xi         |     $\xi$     |     `\xi`     |    $\Xi$   |    `\Xi`   |
| pi         |     $\pi$     |     `\pi`     |    $\Pi$   |    `\pi`   |
| rho        |     $\rho$    |     `\rho`    |     --     |     --     |
| sigma      |    $\sigma$   |    `\sigma`   |  $\Sigma$  |  `\Sigma`  |
| tau        |     $\tau$    |     `\tau`    |     --     |     --     |
| upsilon    |   $\upsilon$  |   `\upsilon`  | $\Upsilon$ | `\Upsilon` |
| phi        |     $\phi$    |     `\phi`    |   $\Phi$   |   `\Phi`   |
| chi        |     $\chi$    |     `\chi`    |     --     |     --     |
| psi        |     $\psi$    |     `\psi`    |   $\Psi$   |   `\Psi`   |
| omega      |    $\omega$   |    `\omega`   |  $\Omega$  |  `\Omega`  |
| varepsilon | $\varepsilon$ | `\varepsilon` |     --     |     --     |

: LaTeX syntax for greek letters. {#tbl-greektable}

```

The last line will adds a caption to the table and `{#tbl-greektable}` creates a label to reference the table in the text using `@tbl-greektable`.

The `gt` function from the `gt` package creates a table from a data frame or R object. Here is an example code to create a table from the first 6 rows of the `mtcars` dataset:

```{r}
#| echo: fenced
#| label: tbl-mtcarsdata
#| eval: true

mtcars |> head() |> 
  gt::gt() |> 
  gt::tab_caption("The MTCARS Dataset")
```

Notice that @tbl-mtcarsdata is easily produced using the `gt()` function with a caption 

Table @tbl-mtcarsdata is referenced by using the label created in the chunk and the `@tbl-mtcarsdata`. To install `gt` run the following line in your console:

```{r}
install.packages("gt")
```

## Citations and Referneces

Qmd documents contains capabilities to add citations and a bibliography. For example, to cite this textbook [@mendenhallSecondCourseStatistics2012], use the `@` symbol followed by a citation identifier from the `.bib` file surrounded by square brackets, `[@mendenhallSecondCourseStatistics2012]`. To cite your textbook again [-@mendenhallSecondCourseStatistics2012] without the authors names, use a `-` sign in front of the `@` symbol, `[-@mendenhallSecondCourseStatistics2012]`. To cite multiple books [@casellaStatisticalInference1990; @rohatgiIntroductionProbabilityStatistics2015; @resnickProbabilityPath2014; @lehmannTheoryPointEstimation1998; @lehmannTestingStatisticalHypotheses2005], add each citation inside the square brackets with the `@` symbol and separate them with semicolons, `[@casellaStatisticalInference1990; @rohatgiIntroductionProbabilityStatistics2015; @resnickProbabilityPath2014; @lehmannTheoryPointEstimation1998; @lehmannTestingStatisticalHypotheses2005]`.

The references will be added automatically at the end of the document. 

In order to use citations and references, the qmd file needs needs a `.bib` file containing all the information of the references. First, save the `.bib` file in the same folder (directory) as your `qmd` file. Then add the line `bibliography: NAME.bib` to the YAML header. Make any changes appropriately to the line, such as the name of the `.bib` file.

### `.bib` File

The `.bib` file is an ordinary text file containing "bib" entries with information about each reference. Below is an example bib entry about R:

```{r}
#| eval: true
#| echo: false

citation() |> toBibtex()
```


Creating a `.bib` file is tedious; however, there are reference managers that can help. I recommend using [Zotero](https://www.zotero.org/), an open-source reference manager designed to import and manage citations. Once a citation is in Zotero, you can export your library as a `.bib` file. Make sure to check your references in Zotero for any mistakes. 

## Math

Quarto is capable of writing mathematical formulas using LaTeX code. A mathematical symbol can be written inline using single `$` signs. For example, `$\alpha$` is viewed as $\alpha$ in a document. To write mathematical formulas on its own line use `$$`. For example, `$$Y=mX+b$$` is viewed as $$Y=mX+b$$.

### Mathematical Notation

| Notation                        | code                              |
|:--------------------------------|:----------------------------------|
| $x=y$                           | `$x=y$`                           |
| $x>y$                           | `$x>y$`                           |
| $x<y$                           | `$x<y$`                           |
| $x\geq y$                       | `$x\geq y$`                       |
| $x\leq y$                       | `$x\leq y$`                       |
| $x^{y}$                         | `$x^{y}$`                         |
| $x_{y}$                         | `$x_{y}$`                         |
| $\bar x$                        | `$\bar x$`                        |
| $\hat x$                        | `$\hat x$`                        |
| $\tilde x$                      | `$\tilde x$`                      |
| $\frac{x}{y}$                   | `$\frac{x}{y}$`                   |
| $\frac{\partial x}{\partial y}$ | `$\frac{\partial x}{\partial y}$` |
| $x\in A$                        | `$x\in A$`                        |
| $x\subset A$                    | `$x\subset A$`                    |
| $x\subseteq A$                  | `$x\subseteq A$`                  |
| $x\cup A$                       | `$x\cup A$`                       |
| $x\cap A$                       | `$x\cap A$`                       |
| $\{1,2,3\}$                     | `$\{1,2,3\}$`                     |
| $\int_a^bf(x)dx$                | `$\int_a^bf(x)dx$`                |
| $\left\{\int_a^bf(x)dx\right\}$ | `$\left\{\int_a^bf(x)dx\right\}$` |
| $\sum^n_{i=1}x_i$               | `$\sum^n_{i=1}x_i$`               |
| $\prod^n_{i=1}x_i$              | `$\prod^n_{i=1}x_i$`              |
| $\lim_{x\to0}f(x)$              | `$\lim_{x\to0}f(x)$`              |
| $X\sim \Gamma(\alpha,\beta)$    | `$X\sim \Gamma(\alpha,\beta)$`    |

: LaTeX syntax for common examples. {#tbl-latextable}

### Greek Letters

| Letter     |   Lowercase   |      Code     |  Uppercase |    Code    |
|:-----------|:-------------:|:-------------:|:----------:|:----------:|
| alpha      |    $\alpha$   |    `\alpha`   |     --     |     --     |
| beta       |    $\beta$    |    `\beta`    |     --     |     --     |
| gamma      |    $\gamma$   |    `\gamma`   |  $\Gamma$  |  `\Gamma`  |
| delta      |    $\delta$   |    `\delta`   |  $\Delta$  |  `\Delta`  |
| epsilon    |   $\epsilon$  |   `\epsilon`  |     --     |     --     |
| zeta       |    $\zeta$    |    `\zeta`    |     --     |     --     |
| eta        |     $\eta$    |     `\eta`    |     --     |     --     |
| theta      |    $\theta$   |    `\theta`   |  $\Theta$  |  `\Theta`  |
| iota       |    $\iota$    |    `\iota`    |     --     |     --     |
| kappa      |    $\kappa$   |    `\kappa`   |     --     |     --     |
| lambda     |   $\lambda$   |   `\lambda`   |  $\Lambda$ |  `\Lambda` |
| mu         |     $\mu$     |     `\mu`     |     --     |     --     |
| nu         |     $\nu$     |     `\nu`     |     --     |     --     |
| xi         |     $\xi$     |     `\xi`     |    $\Xi$   |    `\Xi`   |
| pi         |     $\pi$     |     `\pi`     |    $\Pi$   |    `\pi`   |
| rho        |     $\rho$    |     `\rho`    |     --     |     --     |
| sigma      |    $\sigma$   |    `\sigma`   |  $\Sigma$  |  `\Sigma`  |
| tau        |     $\tau$    |     `\tau`    |     --     |     --     |
| upsilon    |   $\upsilon$  |   `\upsilon`  | $\Upsilon$ | `\Upsilon` |
| phi        |     $\phi$    |     `\phi`    |   $\Phi$   |   `\Phi`   |
| chi        |     $\chi$    |     `\chi`    |     --     |     --     |
| psi        |     $\psi$    |     `\psi`    |   $\Psi$   |   `\Psi`   |
| omega      |    $\omega$   |    `\omega`   |  $\Omega$  |  `\Omega`  |
| varepsilon | $\varepsilon$ | `\varepsilon` |     --     |     --     |

: LaTeX syntax for greek letters. {#tbl-greektable}


## Rendering a Document

A qmd file can be rendered into either an html file, pdf document or word document. Rendering the qmd file to an html file or word document can be easily done using the knit button above. However, rendering the qmd file to a pdf document requires LaTeX to be installed. There are two methods to install LaTeX: from the LaTeX website or from R. I recommend installing the full LaTeX distribution from the <https://www.latex-project.org/get/>. This provides you with everything you may need. You can also install it from R:

```{r, eval=FALSE}
install.packages("tinytex")
tinytex::install_tinytex()
```

You will only need to run these lines of code once and then you can render pdf documents easily.

### HTML {.unnumbered}

```{default}
#| echo: true
---
format: html
---
```


### PDF {.unnumbered}

```{default}
#| echo: true
---
format: pdf
---
```


### Word Document {.unnumbered}

```{default}
#| echo: true
---
format: docx
---
```


## Resources and Tips

### Quarto {.unnumbered}

-   [Quarto](https://quarto.org/)
-   [Markdown Basics](https://quarto.org/docs/authoring/markdown-basics.html)
-   [Formats](https://quarto.org/docs/reference/)

### RMarkdown {.unnumbered}

-   [RStudio](https://rmarkdown.rstudio.com/)
-   [Bookdown](https://bookdown.org/yihui/rmarkdown/)

### YAML {.unnumbered}

-   [UCLA Resource](https://stats.oarc.ucla.edu/stat/data/rmarkdown/rmarkdown_seminar_flat.html#yaml-header)
-   [Reproducible Research](https://monashdatafluency.github.io/r-rep-res/yaml-header.html)
-   [RMarkdown Crash Course](https://zsmith27.github.io/rmarkdown_crash-course/lesson-4-yaml-headers.html)

### Tips {.unnumbered}

-   Render your document often so it easier to identify problems with rendering

-   The Visual Mode in RStudio eases the process of creating a document and makes it more bearable.
-   YAML is tricky with spacing. Make sure that spaces when indenting options. Also make sure that there are not extra spaces at the end of each line.

## References
